Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-30 doc/MS_v4.Rmd
more MS_v4 edits

-old-
Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. The number of crossovers per meiotic cell is the  genome-wide recombination rate. Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two, one for each sex, genome-wide recombination rates which display distinct evolutionary trajectories.

@@

Meiosis converts diploid germ cells into haploid gametes. During meiosis I, DNA crossovers aid the separation of homologous chromosomes by physically linking them and establishing tension between them on the spindle (REF?). The wrong number of recombination events can disrupt chromosomal segregation, leading to infertility, miscarriage, and birth defects (Hassold and Hunt, 2001). Recombination also shapes evolution by shuffling the combinations of genetic variants offspring inherit. Recombination affects the fates of beneficial and deleterious mutations (Felsenstein, 1974; Fisher, 1930; Hill and Robertson, 1966) and interacts with natural selection to leave gradients in genomic patterns of diversity [@begun1992; @nachman2012; @charlesworth1993; @cutter2013; @smith1974] (Begun and Aquadro, 1992; Charlesworth et al., 1993; Cutter and Payseur, 2013; Maynard Smith and Haigh 1974).
The role of recombination in facilitating meiotic chromosome assortment suggests that the total number of crossovers in a cell – the genome-wide recombination rate – is an important cellular characteristic connected to organismal fitness. The dual pressures of ensuring at least one crossover per chromosome and minimizing levels of DNA damage and ectopic exchange are thought to impose lower and upper thresholds on the genome-wide recombination rate (Inoue; Nagaoka et al., 2012). Yet, within these bounds, individuals from the same species can vary substantially in crossover number (Kong et al., 2008; Gruhn et al., 2013; Ma et al., 2015; Johnston et al., 2016).
Sex is perhaps the most notable axis along which recombination rate varies. Broadly speaking, sexual dimorphism in the genome-wide recombination rate assumes two forms. In species such as Drosophila melanogaster, one sex completes meiosis without forming crossovers (“achiasmy”), while the other sex recombines (Haldane, 1922; Huxley, 1928; Burt et al., 1991). Alternatively, in most species with recombination, crossovers occur in both sexes but at different rates (“heterochiasmy”). In these species, females tend to recombine more than males (Bell, 1982; Brandvain and Coop, 2012; Burt et al., 1991; Lenormand and Dutheil, 2005; Lorch, 2005). In plants, heterochiasmy is correlated with the opportunity for haploid selection (Lenormand and Dutheil, 2005). Despite the establishment of these interspecific trends, an understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of variation in female and male recombination rates within species have come from outbred populations of humans (Kong et al., 2004, 2008, 2014; Halldorsson et al., 2019; Gruhn et al., 2013), dogs (Campbell et al. 2014), cattle (Ma et al., 2015; Shen et al., 2018), and Soay sheep (Johnston et al., 2016), in which the role of sex is confounded with the contributions of genetic variation. Although it is known that the level and direction of heterochiasmy can differ among species (Brandvain and Coop 2012; Lenormand and Dutheil, 2005), the correlation between female and male recombination rates among closely related species remains poorly documented. Direct contrasts between the two sexes across a common, diverse set of genomic backgrounds that represent recent timescales would reveal whether the genome-wide recombination rate evolves differently in males and females.
Examining variation in the total number of crossovers in a sex-specific manner could also illuminate evolutionary connections between recombination rate and crossover positioning. Analyses of meiotic chromosome morphology in Arabidopsis thaliana, Caenorhabditis elegans, and Mus musculus suggest that the sex with more recombination usually has longer chromosome axes (Cahoon and Libuda, 2019). A survey of 51 species found conserved sex differences in the recombination landscape, including telomere-biased placement of crossovers in males but not in females (Sardell and Kirkpatrick, 2020). The degree to which a crossover reduces the probability of another crossover nearby (crossover interference) also differs between females and males (Otto and Payseur 2019). 
The house mouse, Mus musculus, provides a compelling system for understanding how sex determines the evolution of recombination. Multiple subspecies share a most recent common ancestor approximately 0.5 million years ago (Geraldes et al. 2011), providing the opportunity to examine natural variation on recent evolutionary timescales. Wild Mus musculus belong to the same species as classical inbred strains of mice, where the molecular and cellular pathways that lead to crossovers have been studied extensively (Baudat et al., 2013; Bolcun-Filas and Schimenti, 2012; Handel and Schimenti, 2010). Single-cell immunofluorescent approaches make it possible to estimate genome-wide recombination rates in individual males and females (Koehler et al. 2002; Peters et al., 1997). A collection of wild-derived inbred strains founded from a variety of geographic locations is available, enabling genetic variation in recombination to be profiled across the species range. Most importantly, by measuring recombination rates in females and males from the same set of wild-derived inbred strains, the evolutionary dynamics of recombination can be directly compared in the two sexes. 

@@

-old--

Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process of a diploid genome producing haploid gametes. The total number of COs at this stage of meiosis is equal to the genome-wide recombination rate (gwRR). Meiosis, and the mixing of haploid gametes, is an important process for evolution, because it can determine a populations’ response to selection by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds [@charlesworth1993]. This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012]. Due to it’s role in facilitating chromosomes disjunction, the genome-wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility. Ensuring at least one crossover per chromosome while minimizing the rates of DNA damage and ectopic recombination are thought to regulate the lower and upper thresholds for a genome-wide recombination rate [@nagaoka2012] (Inoue). Yet, within these thresholds the genome-wide recombination rate still varies across species and individuals.

Sex, or the version of gametogenesis, is one of the most notable axes of this form of variation. There are two classes of sexual dimorphism in recombination rates; the first is called achiasmy and referred to where one sex recombines [@huxley1928; @lenormandDuthiel_2005]. The second, which is the focus of this manuscript, is the instance where both sexes recombine [@lenormandDuthiel_2005]. An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog [@wong2010dog], cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay] petite?  and mouse [@srivastava2017CC] in which the role of sex is confounded with the contributions of genetic variation.

**<evolutionary patterns of sexual dimorphism in traits such as body size, coloration, and – can be traced to sexual antagonism / directional selection due to --- differences in. not sure what I meant here**

For many secondary sexually dimorphic traits, the mode of selection can be traced to the divergent targets of selection. In anisogamy systems where gametogenesis is modified to produce distinct gametes across sexes, the meiotic program is the same and there is no first principles which would predict the evolution of sexually dimorphism in crossover number per cell. Yet heterochiasmy is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Asymmetrical cell division is one of the best distinguishers for sex in metazoans, however there are exceptions, including scarid flies and brown algae [@gorelick2017], yet empirical data on meiotic recombination from such organisms is missing. In the organisms for which there are empirical data on meiotic recombination rates, it is clear that the relationship between female and male recombination rates can differ among species, however comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<!-- 4.	Additional metrics (more nuanced) -->

When considering the evolutionary patterns for genome based metric is to also consider patterns at the individual chromosome scale. The genome-wide recombination rate can be decomposed into single chromosome patterns, referred to as recombination landscapes. There are conserved patterns relevant to heterochiasmy at this scale. Recently two specific features have been highlighted as conserved patterns related heterochiasmy, specifically the recombination landscape. In a review of data from mus musculus, C. elegans and arabidopsis thalania with meiotic chromosome morphology (axis-loop structure) quantified in both sexes, @CahoonLibuda2019 find that the sex with longer chromosome axis generally has higher recombination rates. A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement [@sardell_sex_2020]. These observations raise the question of how conservation of sexually dimorphic genome-wide recombination rates and its’ decomposed traits.

Integrating patterns across multiple scales – into a unified evolutionary pattern requires a good model system with multiple evolutionary levels for gauging the robustness of patterns across multiple scales. The house mouse, (Mus musculus), arose from a recent radiation (0.5 mya to 5 mya) (@geraldes2011; @geraldes2008) into several subspecies providing an opportunity to interrogate natural variation at short evolutionary scales. This short evolutionary scale is valuable for increasing heterochiasmy observations, since recombination rates are known to evolve rapidly (cite). Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway [@baudat2013; @handel2010],  and previous crosses for understanding the genetic architecture of recombination rate variation [@dumont2011; @murdoch2010; @Wang2017island]. Additionally due to their global distribution allows comparisons of from a broad geographic range that might be missed with classical inbred strains. Wild derived inbred strains of house mouse, enable one of the most direct comparison of female and male gametogenesis, since besides the sex chromosomes, the mouse for each genome are identical. Additionally, this model has advantages for single cell cytology approaches, enabling integration of data at a closer connection to the molecular pathway and meiotic program.

In this study, we quantify gwRRs from both sexes, in a panel of inbred mouse strains. We observed highly distinct evolutionary trajectories between males and females. This is due to rapid male specific evolution in the gwRR from two strains in musculus and 1 strain in molossinus. We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale. This pattern of rapid evolution was not observed in females. In our comparisons of recombination landscapes across sex, our results confirm previously reported patterns for sex differences in the general recombination landscape (Sardell and Kirkpatrick, 2020). Intriguingly, these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns.

@@

@@

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` house mice (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`; sex: p = **`r MLH1.MMlmer_results$sex.results`;** subspecies: p = `r MLH1.MMlmer_results$sub.sp`; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval`).

@@

@@

Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD`  (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results` ; M2: p = 
`r sum.Q1.glm_totSC_M2$coefficients[,4][8]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,4][8]` ) and short bivalent length (M1: p = `r Q1.lmer_short_M1_results$sex.results`}; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]/10000`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]/10000`). The existence of some subspecies and strain effects on total SC and short bivalent SC Length further indicates that SC length has evolved among strains and among subspecies (See Supplemental Tables 5 and 8). In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males recombine more. This pattern implies that in high-recombination strains, spermatocytes have less space than oocytes in which to position additional crossovers.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; paired t-test; p = `r Q1.strain.F1.Pos$p.value`). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results`}; M2: p =  `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p =  `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

@@

>>>>

@@

Meiosis converts diploid germ cells into haploid gametes. During meiosis I, DNA crossovers aid the separation of homologous chromosomes by physically linking them and establishing tension between them on the spindle **(REF?)**. The wrong number of recombination events can disrupt chromosomal segregation, leading to infertility, miscarriage, and birth defects [@hassold2001]. Recombination also shapes evolution by shuffling the combinations of genetic variants offspring inherit. Recombination affects the fates of beneficial and deleterious mutations [@felsenstein1974; @fisher1930; @hill1966] and interacts with natural selection to leave gradients in genomic patterns of diversity [@begun1992; @nachman2012; @charlesworth1993; @cutter2013; @smith1974].

The role of recombination in facilitating meiotic chromosome assortment suggests that the total number of crossovers in a cell – the genome-wide recombination rate – is an important cellular characteristic connected to organismal fitness. The dual pressures of ensuring at least one crossover per chromosome and minimizing levels of DNA damage and ectopic exchange are thought to impose lower and upper thresholds on the genome-wide recombination rate [@nagaoka2012; @inoue2002]. Yet, within these bounds, individuals from the same species can vary substantially in crossover number [@Kong2008; @gruhn2013; @ma2015_cattle; @johnston2016_soay].

Sex is perhaps the most notable axis along which recombination rate varies. Broadly speaking, sexual dimorphism in the genome-wide recombination rate assumes two forms. In species such as Drosophila melanogaster, one sex completes meiosis without forming crossovers (“achiasmy”), while the other sex recombines [@huxley1928; @haldane1922sex; @burt1991]. Alternatively, in most species with recombination, crossovers occur in both sexes but at different rates (“heterochiasmy”). In these species, females tend to recombine more than males (Bell, 1982; Brandvain and Coop, 2012; Burt et al., 1991; Lenormand and Dutheil, 2005; Lorch, 2005). [@lorch2005; @brandvain2012scrambling; @burt1991; @lenormandDuthiel_2005 ] (**can't find Bell 1982 citation**). In plants, heterochiasmy is correlated with the opportunity for haploid selection [@lenormandDuthiel_2005]. Despite the establishment of these interspecific trends, an understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of variation in female and male recombination rates within species have come from outbred populations of humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog [@wong2010dog], cattle [@ma2015_cattle; @Shen2018_cattle],  sheep [@johnston2016_soay] and mouse [@srivastava2017CC] (Kong et al., 2004, 2008, 2014; Halldorsson et al., 2019; Gruhn et al., 2013), dogs (Campbell et al. 2014), cattle (Ma et al., 2015; Shen et al., 2018), and Soay sheep (Johnston et al., 2016), in which the role of sex is confounded with the contributions of genetic variation. Although it is known that the level and direction of heterochiasmy can differ among species [@brandvain2012scrambling; @lenormandDuthiel_2005], the correlation between female and male recombination rates among closely related species remains poorly documented. Direct contrasts between the two sexes across a common, diverse set of genomic backgrounds that represent recent timescales would reveal whether the genome-wide recombination rate evolves differently in males and females.
Examining variation in the total number of crossovers in a sex-specific manner could also illuminate evolutionary connections between recombination rate and crossover positioning. Analyses of meiotic chromosome morphology in Arabidopsis thaliana, Caenorhabditis elegans, and Mus musculus suggest that the sex with more recombination usually has longer chromosome axes [@CahoonLibuda2019]. A survey of 51 species found conserved sex differences in the recombination landscape, including telomere-biased placement of crossovers in males but not in females [@sardell_sex_2020]. The degree to which a crossover reduces the probability of another crossover nearby (crossover interference) also differs between females and males [@ottoPaysuer2019]. 
The house mouse, Mus musculus, provides a compelling system for understanding how sex determines the evolution of recombination. Multiple subspecies share a most recent common ancestor approximately 0.5 million years ago [@geraldes2011], providing the opportunity to examine natural variation on recent evolutionary timescales. Wild Mus musculus belong to the same species as classical inbred strains of mice, where the molecular and cellular pathways that lead to crossovers have been studied extensively [@baudat2013; @handel2010; @bolcun2012]. Single-cell immunofluorescent approaches make it possible to estimate genome-wide recombination rates in individual males and females [@koehler2002; @peters_1997]. A collection of wild-derived inbred strains founded from a variety of geographic locations is available, enabling genetic variation in recombination to be profiled across the species range. Most importantly, by measuring recombination rates in females and males from the same set of wild-derived inbred strains, the evolutionary dynamics of recombination can be directly compared in the two sexes.

@@

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` -- 14 wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ). Mean MLH1 focus counts for `r sum(main.table$Nmice)` -- 188 mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` -- 21.77 spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` -- 1,742 spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` -- 17.85 oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` -- 1,427 oocytes). **<insert statement for cell sample size number>**

Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the difference in counts between the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains (*musculus^PWD^* and _molossinus^MSM^_) exhibit higher counts in males. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23) of foci (Figure 1C). Strain mean MLH1 focus counts from females and males are uncorrelated (Spearman’s $\rho$ = `r sperman.corr.FM` -- 0.08; p = `r sperman.corr.FM.pval$p.value` -- 0.84) across the set of strains.

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` -- 137 house mice (from three subspecies; _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_) (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`-- 10-4; sex: p = `r MLH1.MMlmer_results$sex.results` -- 3.64 x 10-6; subspecies: p = `r MLH1.MMlmer_results$sub.sp` -- 9.69 x 10-4; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval` -- 1.8 x 10-4).

The effect of subspecies is no longer significant in a model treating all factors as fixed effects (M2; musculus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]` -- 0.24, molossinus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]` -- 0.1), highlighting strain and sex as salient variables. Two strains exhibit strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` -- 1.78 x 10-6; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]` -- 0.02), with sex-strain interactions involving three strains (M3; *domesticus^G^* p < x 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p < x 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` -- 3.87 x 10-4).

In separate analyses of males (M4; n = `r length(male_mouse.avs_for.analysis$mouse)` -- 71), three strains disproportionately shape MLH1 focus count (as observed in Figure 1C): *musculus^PWD^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][7]` -- 3.6 x 10-7; effect = `r sum_MLH1.male.glm.M4$coefficients[7]` -- 6.11 foci **(add CI’s here)**, *molossinus^MSM^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]` -- 6.3 x 10-9; effect = `r sum_MLH1.male.glm.M4$coefficients[8]`-- 6.91), and *musculus^SKIVE^* (p = -- 8.22 x 10-4; effect = -- 4.04). These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. In females (M4; n= `r length(female_mouse.avs_for.analysis$mouse)` --76), three strains affect MLH1 focus count: *domesticus^G^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][2]` -- 8.7 x 10-6; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[2]` -- 3.3), *molossinus^MSM^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][5]` -- 2.43 x 10-5; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[5]` -- 2.99), and *domesticus^LEW^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][3]`  -- 0.03; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[3]` -- 1.69). Strain effect sizes in females are modest in magnitude compared to those in males. Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

-old-

@@

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` house mice (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`; sex: p = `r MLH1.MMlmer_results$sex.results`; subspecies: p = `r MLH1.MMlmer_results$sub.sp`; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval`).

@@

The variation in sex differences in recombination we discovered provided an opportunity to determine whether sex differences in chromatin compaction, as measured by the length of the synaptonemal complex (SC), are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (t-tests; all p < 0.05, except short bivalents in _musculus^SKIVE^_, p = `r Q1_short.SC_t.test.SKIVE$p.value` -- 0.11).
Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD` -- 1.26 (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` -- 1.52 (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results` -- 2.56 x 10-31; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,4][8]` -- 2.56 x 10-8; M3: p =  `r sum.Q1.glm_totSC_M3$coefficients[,4][8]` -- 2.56 x 10-8) and short bivalent SC length (M1: p = `r Q1.lmer_short_M1_results$sex.results` -- 1.12 x 10-11; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]/10000` -- 1.33 x 10-7; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]/10000` -- 1.33 x 10-7). The existence of some subspecies and strain effects on total SC length and short bivalent SC length (Supplemental Tables 5 and 8) further indicates that SC length has evolved among strains and among subspecies. 
In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males recombine more. This pattern implies that in high-recombination strains, spermatocytes have less space than oocytes in which to position additional crossovers.

-old-

@@

Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD`  (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results` ; M2: p = 
`r sum.Q1.glm_totSC_M2$coefficients[,4][8]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,4][8]` ) and short bivalent length (M1: p = `r Q1.lmer_short_M1_results$sex.results`; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]/10000`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]/10000`). The existence of some subspecies and strain effects on total SC and short bivalent SC Length further indicates that SC length has evolved among strains and among subspecies (See Supplemental Tables 5 and 8). In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males recombine more. This pattern implies that in high-recombination strains, spermatocytes have less space than oocytes in which to position additional crossovers.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )` -- 0.68 **add s.d here?**; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )` --  0.56; paired t-test; p = `r Q1.strain.F1.Pos$p.value` -- 8.49 x 10-4). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results` -- 2.82 x 10-26; M2: p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]` -- 3.96 x 10-8; M3: p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]` -- 3.96 x 10-8).

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (p = `r Q1.IFD.PER_t.test_KAZ$p.value` -- 0.33). Examination of IFD~norm~ distributions indicates that females are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Models treating IFD~norm~ as the dependent variable support the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results` -- 9.08 x 10-12; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]` -- 0.01; M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]` -- 0.01). In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) (Supplemental Table 13 and 14) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers **(Figure 4C and Supplemental Figure X)** further suggests that interference distances vary more in females than in males, and that males display a stronger telomeric bias in the placement of the distal crossover.

In summary, controlling for differences in SC length (chromatin compaction) indicates that interference is consistently stronger in males, whereas interference on the physical scale is similar in the two sexes.

-old-

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; paired t-test; p = `r Q1.strain.F1.Pos$p.value`). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results`; M2: p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

@@

Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")`-- 9,569) have either one or two MLH foci (Supplemental Figure 2). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Supplemental Figure 2). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` -- 0.33 in _musculus^SKIVE^_, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` -- 0.44 in _musculus^PWD^_, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` -- 0.51 in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _molossinus^MSM^_: p = `r PWD.MSM.2CO.prop.test$p.value` -- 0.37; _musculus^PWD^_ vs. *musculus^KAZ^*: p = `r PWD.KAZ.2CO.prop.test$p.value` -- 1.23 x 10-31; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = `r MSM.MOLF.2CO.prop.test$p.value` -- 2.34 x  10-6). These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number across multiple bivalents.

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` -- 76 early zygotene and `r sum( (DMC1table.main$ncells.y) )` -- 75 late zygotene spermatocytes from two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; -- `r 1/1000000`p < 10-6). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval` -- 0.66). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.
